Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2022-02-22 Manuscript/03-Simulation.Rmd
Revise Simulation Prediction Results (#23)

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]\bg{, linear lasso model and spikeSlabGAM [TODO: add citation here].[TODO: explain about linear lasso model and spikeSlabGAM model]}. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for nonlinear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence use the default choices. Both mgcv and proposed models allow customized smoothing functions and we choose the cubic regression spline. We controll the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. We use 5-fold CV with the default selection criteria to select the final model for COSSO models, SB-GAM and the proposed models. 20 default candidates of tuning parameters ($s_0$ in BHAM, $\lambda_0$ in SB-GAM) are examined for SB-GAM and the proposed models which allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].

@@

To evaluate the predictive performance of the models, the statistics, $R^2$ for Gaussian model and AUC for binomial model calculated based on the testing dataset, are averaged <!--and deviance are calculated for Poisson model,--> across 50 simulations. \bg{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.}
$$
\bg{MCC = \frac{TP\times TN - FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}},}
$$
\bg{where TP, TN, FP, and FN are true positives, true negatives, false positives, and false negatives, respectively.}
\subsection{\bg{Simulation Results}}
\subsubsection{\bg{Prediction Performance}}
The predictive performances have a consistent pattern across the two distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv doesn't support high-dimensional analysis, i.e. the number of coefficients are greater than the sample size, and hence not evaluated when $p=100, 200$. mgcv predicts well when $p$ is small or moderate ($p = 4, 10$), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS performs as good as mgcv if not better. However, in high dimensional case where we mimic the situation the signals are extremely sparse, SB-GAM has better performance than the proposed method. However, the BHAM-CD has extreme computational advantage over SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\begin{itemize}
\item \bg{When the linear assumptions are violated, the linear lasso model performs extremely bad. adaptive cosso have improvement upon cosso.}
\item \bg{To note, the drastic deteriation in the prediction of mgcv is caused by the model would not converge in those scenarios. A null model were returned as a final model. }
\item \bg{Within the simulations where the functional form of the predictors are linear, we see that the proposed model similar to the lasso model regardless of the type outcome, and better moderately better than the model of comparison. (See Table Lnr Gause, and supporting table Lnr-binom.) Particularly, in gaussina simulations, the proposed model shows great advantage compared to SB-GAM (numeralize the advantage), and slightly improvement compare to ssGAM (numeralize the advantage). This shows that the model can be applied without constraining to the assumption where all functional form are non-linear.}
\end{itemize}

>>>>

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]\bg{, linear lasso model and spikeSlabGAM [TODO: add citation here].[TODO: explain about linear lasso model and spikeSlabGAM model]}. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for nonlinear effect interpolation and prediction. \bg{mgcv doesn't support analyses where the number of parameters exceeds the sample size.} SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence use the default choices. Both mgcv and proposed models allow customized smoothing functions and we choose the cubic regression spline. We controll the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. We use 5-fold CV with the default selection criteria to select the final model for COSSO models, SB-GAM and the proposed models. 20 default candidates of tuning parameters ($s_0$ in BHAM, $\lambda_0$ in SB-GAM) are examined for SB-GAM and the proposed models which allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].

@@

To evaluate the predictive performance of the models, the statistics, $R^2$ for Gaussian model and AUC for binomial model calculated based on the testing dataset, are averaged <!--and deviance are calculated for Poisson model,--> across 50 simulations. \bg{To evaluate the variable selection performance of the models, we record the set of variables each model selects and calculate the averaged false discover rate (FDR), false negative rate(FNR), and Matthews correlation coefficient (MCC). For method that doesn't automatically achieve variable selection, we set alpha level at 0.05 for mgcv that relies on hypothesis testing for variable selection, and soft-threshold at 0.5 for spikeSlabGAM given the marginal inclusion probabilities. For the two methods, BHAM and spikeSlabGAM, that are capable of bi-level selection, we record the probability that the linear and nonlinear components of each predictors are selected in the models.}
$$
\bg{MCC = \frac{TP\times TN - FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)}},}
$$
\bg{where TP, TN, FP, and FN are true positives, true negatives, false positives, and false negatives, respectively.}
\subsection{\bg{Simulation Results}}
\subsubsection{\bg{Prediction Performance}}
\bg{Among the main set of simulations where the functional forms of the predictors are mostly non-linear,} the predictive performances have a consistent pattern across the two distributions of outcomes. (See Tables \ref{tab:gaus} and \ref{tab:bin_auc}) \bg{The proposed model, BHAM, predicts as good as, if not better than, other high dimensional spline models. Specifically, BHAM shows greate improvement over COSSO methods [TODO: add statsitics] and the spikeSlabGAM model [TODO: add statsitics]. When comparing to SB-GAM, BHAM performs better in lower dimensional cases ($p=4,10$), and equally good or slightly worse [TODO: re-evaluate after tuning the models] in high-dimensional cases ($p=50,100, 200$). As previously hypothesized, the linear lasso model predicts less accurate compared to other flexibly models across all scenario; mgcv performs extremely well in low-dimensional case ($p = 4, 10$), and deteriorates as the dimensionality increases until not applicable. To note, mgcv still fit models with increasing in large dimension case ($p=50$). But the algorithm fails to converge within the default number of iterations, and hence leads to bad prediction performance.} In high dimensional cases where we mimic the situation the signals are extremely sparse, SB-GAM has better performance than the proposed method. However, the BHAM-CD has extreme computational advantage over SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\bg{We also examine the prediction performance when the functional form of predictors are linear. (See support information Table 1, and 2 [TODO: add hyperlink to the github repo]) The proposed model, BHAM, has similar performance as the linear lasso model regardless of the type outcome. This observation demonstrates that BHAM is a flexible model, and has good prediction performance regardless the underlying functional form of predictors. spikeSlabGAM have similar prediction performance to BHAM (numeralize the advantage). Surprisingly, SB-GAM doesn't perform well in high-dimensional Gaussian outcome scenarios. (numeralize the advantage)}

### 2022-02-22 Manuscript/SS_GAM.tex
Revise Simulation Prediction Results (#23)

In this section, we compare the performance of the proposed models to
four alternative models: component selection and smoothing operator
(COSSO) \citep{Zhang2006GAM}, adaptive COSSO \citep{Storlie2011},
generalized additive models with automatic smoothing \citep{Wood2011},
SB-GAM
\citep{Bai2021}\textcolor{red}{, linear lasso model and spikeSlabGAM [TODO: add citation here].[TODO: explain about linear lasso model and spikeSlabGAM model]}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component are different for extra flexibility. Generalized additive
models with automatic smoothing, hereafter \textit{mgcv}, is one of the
most popular models for nonlinear effect interpolation and prediction.
SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and
adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive
models with automatic smoothing with R package \texttt{mgcv 1.8-31},
SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do
not provide flexibility to define smoothing functions, and hence use the
default choices. Both mgcv and proposed models allow customized
smoothing functions and we choose the cubic regression spline. We
controll the dimensionality of each smoothing function, 10 bases, for
all different choices of smoothing functions. We use 5-fold CV with the
default selection criteria to select the final model for COSSO models,
SB-GAM and the proposed models. 20 default candidates of tuning
parameters (\(s_0\) in BHAM, \(\lambda_0\) in SB-GAM) are examined for
SB-GAM and the proposed models which allow user-specification of tuning
candidates. All computation was conducted on a high-performance 64-bit
Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680
processors and 24G of RAM per core and R 3.6.2 \citep{R}.

@@

The predictive performances have a consistent pattern across the two
distributions of outcomes. Across all the scenarios, COSSO and adaptive
COSSO have the least favorable performance among the applicable methods
examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv
doesn't support high-dimensional analysis, i.e.~the number of
coefficients are greater than the sample size, and hence not evaluated
when \(p=100, 200\). mgcv predicts well when \(p\) is small or moderate
(\(p = 4, 10\)), and deteriorate when the number of predictors increase.
Among the three fast-computing Bayesian hierarchical models, the
proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when
the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS
performs as good as mgcv if not better. However, in high dimensional
case where we mimic the situation the signals are extremely sparse,
SB-GAM has better performance than the proposed method. However, the
BHAM-CD has extreme computational advantage over SB-GAM (see Table
\ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

\begin{itemize}
\item \textcolor{red}{When the linear assumptions are violated, the linear lasso model performs extremely bad. adaptive cosso have improvement upon cosso.}
\item \textcolor{red}{To note, the drastic deteriation in the prediction of mgcv is caused by the model would not converge in those scenarios. A null model were returned as a final model. }
\item \textcolor{red}{Within the simulations where the functional form of the predictors are linear, we see that the proposed model similar to the lasso model regardless of the type outcome, and better moderately better than the model of comparison. (See Table Lnr Gause, and supporting table Lnr-binom.) Particularly, in gaussina simulations, the proposed model shows great advantage compared to SB-GAM (numeralize the advantage), and slightly improvement compare to ssGAM (numeralize the advantage). This shows that the model can be applied without constraining to the assumption where all functional form are non-linear.}
\end{itemize}

>>>>

In this section, we compare the performance of the proposed models to
four alternative models: component selection and smoothing operator
(COSSO) \citep{Zhang2006GAM}, adaptive COSSO \citep{Storlie2011},
generalized additive models with automatic smoothing \citep{Wood2011},
SB-GAM
\citep{Bai2021}\textcolor{red}{, linear lasso model and spikeSlabGAM [TODO: add citation here].[TODO: explain about linear lasso model and spikeSlabGAM model]}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component are different for extra flexibility. Generalized additive
models with automatic smoothing, hereafter \textit{mgcv}, is one of the
most popular models for nonlinear effect interpolation and prediction.
\textcolor{red}{mgcv doesn't support analyses where the number of parameters exceeds the sample size.}
SB-GAM is the first spike-and-slab lasso GAM. We implement COSSO and
adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive
models with automatic smoothing with R package \texttt{mgcv 1.8-31},
SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do
not provide flexibility to define smoothing functions, and hence use the
default choices. Both mgcv and proposed models allow customized
smoothing functions and we choose the cubic regression spline. We
controll the dimensionality of each smoothing function, 10 bases, for
all different choices of smoothing functions. We use 5-fold CV with the
default selection criteria to select the final model for COSSO models,
SB-GAM and the proposed models. 20 default candidates of tuning
parameters (\(s_0\) in BHAM, \(\lambda_0\) in SB-GAM) are examined for
SB-GAM and the proposed models which allow user-specification of tuning
candidates. All computation was conducted on a high-performance 64-bit
Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680
processors and 24G of RAM per core and R 3.6.2 \citep{R}.

@@

\textcolor{red}{Among the main set of simulations where the functional forms of the predictors are mostly non-linear,}
the predictive performances have a consistent pattern across the two
distributions of outcomes. (See Tables \ref{tab:gaus} and
\ref{tab:bin_auc})
\textcolor{red}{The proposed model, BHAM, predicts as good as, if not better than, other high dimensional spline models. Specifically, BHAM shows greate improvement over COSSO methods [TODO: add statsitics] and the spikeSlabGAM model [TODO: add statsitics]. When comparing to SB-GAM, BHAM performs better in lower dimensional cases ($p=4,10$), and equally good or slightly worse [TODO: re-evaluate after tuning the models] in high-dimensional cases ($p=50,100, 200$). As previously hypothesized, the linear lasso model predicts less accurate compared to other flexibly models across all scenario; mgcv performs extremely well in low-dimensional case ($p = 4, 10$), and deteriorates as the dimensionality increases until not applicable. To note, mgcv still fit models with increasing in large dimension case ($p=50$). But the algorithm fails to converge within the default number of iterations, and hence leads to bad prediction performance.}
In high dimensional cases where we mimic the situation the signals are
extremely sparse, SB-GAM has better performance than the proposed
method. However, the BHAM-CD has extreme computational advantage over
SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the
prediction accuracy.

\textcolor{red}{
We also examine the prediction performance when the functional form of predictors are linear. (See support information Table 1, and 2 [TODO: add hyperlink to the github repo]) The proposed model, BHAM, has similar performance as the linear lasso model regardless of the type outcome. This observation demonstrates that BHAM is a flexible model, and has good prediction performance regardless the underlying functional form of predictors. spikeSlabGAM have similar prediction performance to BHAM (numeralize the advantage). Surprisingly, SB-GAM doesn't perform well in high-dimensional Gaussian outcome scenarios. (numeralize the advantage)}
